Live Functional Programming with Typed Holes

### 2018-02-09 abstract.tex
linebreak issue
rough notes for examples section

\emph{Live programming environments} aim to provide programmers (and, in some
cases, audiences) with continuous feedback about a program's dynamic behavior
throughout the editing process.
%
Unfortunately, because programming languages typically assign dynamic meaning
only to programs that are \emph{complete}---syntactically well-formed and free
of static type errors---live programming features are often severely restricted,
if available at all, when programs are incomplete.
%
Incomplete edit states are frequently encountered when writing programs, so they
are, arguably, when live programming services would be most valuable.
%
Defining a semantics for incomplete functional programs would help bridge the
gap between well-formed states and allowing live programming environments to
provide continuous, uninterrupted, and meaningful feedback.

@@

Formally, our semantics combines features from gradual type theory (to handle
incomplete types) and contextual modal type theory (CMTT, which provides a
logical foundation for hole environment tracking) with novel structure necessary
to prove key metatheoretic properties, including (1) a type safety property that
accounts for expression holes (i.e. free metavariables, from the perspective of
CMTT), and (2) commutativity and confluence properties that guarantee that the
``edit-and-resume'' feature is sound.
%
We have mechanized our formal development using the Agda proof assistant.
%
We additionally describe a simple reference implementation, called Hazelnut
Live, that exposes a language of structured edit actions that insert holes as
necessary to guarantee that every edit state has some (possibly incomplete)
type, based on the edit action calculus in the prior work by \citet{popl-paper}.
%
Taken together, the result is a proof-of-concept live functional programming
environment where dynamic feedback is truly continuous, i.e. it is available for
every possible edit state. % i.e. \emph{every edit state is guaranteed to have
non-trivial static and dynamic meaning}.

>>>>

\emph{Live programming environments} aim to provide programmers (and, in some
cases, audiences) with continuous feedback about a program's dynamic behavior
throughout the editing process.
%
Unfortunately, because programming languages typically assign dynamic meaning
only to programs that are \emph{complete}---syntactically well-formed and free
of static type errors---live programming features are often severely restricted,
if available at all, when programs are incomplete.
%
Incomplete edit states are frequently encountered when writing programs, so they
are, arguably, when live programming services would be most valuable.
%
Defining a semantics for incomplete functional programs would help bridge the
gap between well-formed states and allow live programming environments to
provide continuous, uninterrupted, and meaningful feedback.

@@

Formally, our semantics combines features from gradual type theory (to handle
incomplete types) and contextual modal type theory (CMTT, which provides a
logical foundation for hole environment tracking) with novel structure necessary
to prove key metatheoretic properties, including (1) a type safety property that
accounts for expression holes (i.e. free metavariables, from the perspective of
CMTT), and (2) commutativity and confluence properties that guarantee that the
``edit-and-resume'' feature is sound.
%
We have mechanized our formal development using the Agda proof assistant.
%
We additionally describe a simple reference implementation, called Hazelnut
Live, that exposes a language of structured edit actions that insert holes as
necessary to guarantee that every edit state has some (possibly incomplete)
type, based on the edit action calculus in the prior work by \citet{popl-paper}.
%
Taken together, the result is a proof-of-concept live functional programming
environment where dynamic feedback is truly continuous, i.e. it is available for
every possible edit state.
% i.e. \emph{every edit state is guaranteed to have
% non-trivial static and dynamic meaning}.

### 2018-02-09 examples.tex
rough notes for examples section

\todo{examples}

>>>>

We introduce \HazelnutLive{} in this section using several examples,
demonstrating how running incomplete programs---with missing expressions and/or
type-inconsistent expressions---can be useful.
%
We will also demonstrate how running even complete programs marked with holes is
useful as a form of debugging.

\parahead{Live Programming with Incomplete Expressions}

\emph{Example 1: Computing Weighted Averages.}
%
type for grades.
%
all student data entered in a list of records.

\begin{lstlisting}
type student_rec =
  { name: string,
    hw1: float, hw2: float, hw3: float,
    midterm: float, final: float };

let grades : list(student_rec) =
  [ { name: "Alice", hw1: 76.0, ... }
  , { name: "Bob", hw1: 88.0, ... }
  , ...
  ];

let weighted_average(g: student_rec) =
  ??;

let weighted_averages = List.map weighted_average grades;
\end{lstlisting}

end of the term, now time to assign grades.
%
weighted average of scores.
%
start by weighting hw1 to be 10\%. the rest of the expression is a hole, but can
run the expression, and see that the weighted averages are too large
%

\begin{lstlisting}
let weighted_average(g: student_rec) =
  (10.0 *. g.hw1) + . ??;
\end{lstlisting}

oops, should be using numbers in the range 0 to 1 rather than 1 to 100 for
percentage points.
%
by seeing the output as the expression is being written, this error is caught
early and avoids the same mistake being made several times.

finish the weights, and get the weighted averages.

\emph{Example 2: Assigning Letter Grades.}
%
next, the teacher must map weighted averages to letter grades.
%
using the following cutoffs type.

\begin{lstlisting}
type grade_cutoffs =
  { a: float, b: float, c: float, d: float, f: float };

let cutoffs =
  { a = ??, b = ??, c = ??, d = ??, f = ?? };
\end{lstlisting}

initially all holes, because it will be different year-to-year based on the
data, differences in course difficulty, and to satisfy fairness criteria.
%
write some code that does the mapping.

\begin{lstlisting}
let letter_grade(n: weighted_average) =
  if n >= cutoffs.a then "A" else
  if n >= cutoffs.b then "B" else
  if n >= cutoffs.c then "C" else
  if n >= cutoffs.d then "D" else
  if n >= cutoffs.f then "F" else
  "Incomplete";

let sorted_weighted_averages = List.sort weighted_averages;

let letter_grades = List.map letter_grade sorted_weighted_averages;
\end{lstlisting}

since each constant in the multi-way if is a hole, none of the weighted averages
resolve to a letter.
%
to start figuring out the constants, first sort the weighted averages.
%
by looking at the data, there is a natural gap between 92 and 89.
%
so, choose to to use 92 as the cutoff.
%
see XXX A's pop out in the lettergrades list.
%
before assigning other cutoffs, want to get a sense if whether this is even a
good choice for A.
%
jump to writing a distribution function.

\begin{lstlisting}
let distribution(list: list(float)) =
  let n = List.length letter_grades in
  List.map
    (\x -> (x, showPercentage (List.length (List.filter ((==) x) list) /. n)))
    ["A","B","C","D","F","Incomplete"];
\end{lstlisting}

this shows that the percentage of As is XXX\%, which seems too small.
%
look back at the sorted data, and see a cluster around 89, and then another gap
between 88 and 85.
%
so, tweak A to be 88.
%
the percentage is updated to XXX\%.

now, continue down the list of sorted averages, picking the next cutoff.
%
this series of edits fills one hole, then another, then another.
%
so after each, the suspended computation from before can simply be resumed.
%
in cases where a previous cutoff is modified (which is no longer marked by any
hole), our semantics provides no caching.
%
however, typical incremental computation can be incorporated to determine which
situations such changes lead to partial reuse of previous computation.
%
we do not model this in our core calculus, since our focus is incremental
computation for the specific new mechanism we study, namely, evaluating around
expression holes.

throughout this example, we described a workflow where the programmers goes back
and forth between different parts of the program.
%
this is only possible to do smoothly because of our treatment of expression
holes.
%
treating any one of them as an exception, as is the common ad-hoc approach to
expression holes, prevents the rest of the useful computation from proceeding
and being reused when that hole is filled.

\parahead{Live Programming with Type Errors}

Understanding and debugging static type errors is notoriously difficult,
particularly for novices.
%
A variety of approaches have been
proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
and explain type errors.
%
One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
that demonstrates a run-time failure, and then displays a compressed execution
trace to the user as a graph.
%
By running type-incorrect programs as far as possible (in particular, until they
would go wrong), our semantics can provide a similar benefit operationally.
%
\rkc{they define a partial evaluation with a form of holes. compare.}

\emph{Example 3: Sum List.}
%
Consider the following buggy program of numbers reported by \citet{XXX} that
attempts to sum a list numbers.
%
The problem is that the base case produces a list rather than a number (in
particular, 0).

\begin{lstlisting}
sumList : list(int) -> int
sumList [] = ?[]?
sumList (n:ns) = n + sumList ns
\end{lstlisting}

\noindent
%
Because the list expression on line 2 does not have time \li{int} as required,
it is wrapped in a (non-empty) hole.
%
Rather than trying to debug the error based on the static error, the programmer
may wish to run a call to the function, say \li{sumList(2)}, in which case
\HazelnutLive{} runs and produces the indeterminate value \li{1 + 2 + ?[]?}
(\cf{} the trace displayed by \citet{Seidel2016}).
%
By observing that the hole expression is being added to the number \li{2}, he
realizes that it needs to be a number, specifically, \li{0}.

%% TODO fold error from Erwig paper.
%% %
%% see that final call on stack does have the right answer, but
%% it's wrapped in a singleton list when the expected type is not
%% a list.
%% %
%% fix is to remove the list, the rest of the computation remains
%% the same, but b/c they were all wrapped in holes, need to re-run.
%% %
%% (add some mechanism for type-consistent non-empty-holes...)

\parasection{Example 4: Stutter}
%
As another example, consider the following function which attempts to produce a
list where every element is repeated twice (borrowed from \citep{XXX}).
%
The combiner function to \li{List.foldr} needs to produce a \li{list(int)}, but
it produces a \li{list(list(int)} instead.

\begin{lstlisting}
stutter : list(int) -> list(int)
stutter xs = List.foldr (\x acc -> ?([x,x] : acc)?) [] xs
\end{lstlisting}

\noindent
%
The editor has a choice about which expression to ``blame'' for the error; the
entire application that forms the body of the lambda is analyzed against the
return type \li{list(int)}, so that is a reasonable choice for the editor to
make; another would be to assume that the arguments \li{[x,x]} and \li{acc} are
both as intended and that only the function \li{(:)} is type-consistent.
%
Although one could imagine a setting in which a user would perform this
reasoning, let's assume the simplest approach for marking consistencies that
wraps the entire application.
%
Running this on \li{stutter [1,2,3]} produces the indeterminate value
\li{?  [1,1] : (? [2,2] : (? [3:3] : [] ?) ?) ?}, which shows the unfolding of
\li{List.foldr}.
%
Inside the innermost hole, the value is shown to be \li{[[3,3]]}; the user
realizes that there are too many levels of nesting.
%
He tries append in place of cons, which addresses the type inconsistency and
produces the desired result.

\parahead{Live Programming for Debugging}

\rkc{not sure about the quicksort example yet}

%% \begin{lstlisting}
%% quicksort : list('a) -> list('a)
%% quicksort [] = []
%% quicksort (x:xs) =
%%   let (left, right) = partition ((<) x) xs in
%%   let (left', right') = (quicksort left, quicksort right) in
%%   left' @ [x] @ right
%% \end{lstlisting}
